%% Submissions for peer-review must enable line-numbering
%% using the lineno option in the \documentclass command.
%%
%% Preprints and camera-ready submissions do not need
%% line numbers, and should have this option removed.
%%
%% Please note that the line numbering option requires
%% version 1.1 or newer of the wlpeerj.cls file, and
%% the corresponding author info requires v1.2

\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions
% \documentclass[fleqn,10pt]{wlpeerj} % for preprint submissions

\newcommand{\sjb}[1]{\textcolor{blue}{[sjb:#1]}}
\newcommand{\cjw}[1]{\textcolor{green}{[cjw:#1]}}

% Simon utils
\newcommand{\be}{\begin{enumerate}}
\newcommand{\ee}{\end{enumerate}}
\newcounter{saveenumi}
\newcommand{\seti}{\setcounter{saveenumi}{\value{enumi}}}
\newcommand{\conti}{\setcounter{enumi}{\value{saveenumi}}}


\title{Conda-Forge: a community driven software packaging ecosystem}

% Will fill in via the magic of jinja2 at some point
\author[1]{First Author}
\author[2]{Second Author}
\affil[1]{Address of first author}
\affil[2]{Address of second author}

% \keywords{Keyword1, Keyword2, Keyword3}

\begin{abstract}
Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}


\section*{Sketch}
\begin{enumerate}
\item Conda-Forge is a binary distribution using conda aimed at making code
  available to users across platforms, operating systems, and languages.
\item Three perspectives:
\begin{enumerate}
\item Users: Use CF via `conda install xyz -c conda-forge`. CF focuses on making
  sure that packages installed from CF are self consistent.
\item Package Maintainers: The CF workflow focuses on using the standard GitHub
  fork, branch, and pull request model. New packages are created via PRs into
  staged-recipes. Updates to existing packages are made by PRs into the
  associated feedstock.
\item Ecosystem Maintainers: Automated tools for keeping the ecosystem
  self-consistent and up to date. Unified build tools across all workers ensures
  package compatibility.
\end{enumerate}
\item Future work will focus on moving to new architectures (arm, PowerPC,
web assembly) and reducing maintainer friction.
\end{enumerate}


\section*{Introduction}
Software is a critical part of modern life, from health to commerce to science.
It is increasingly important to be able to distribute and share software.
Distribution is particularly important in the context of open-source community
developed software, where code is readily available, but it is not always easy
to put pieces together. In scientific software, many groups write software for
their own use, but want to share the code more broadly. Scientific software is
frequently difficult to get working, because the need for speed typically
requires complex dependencies using compiled languages.

Packaging software and uploading to package repositories such as PyPI~\cite{??}
and anaconda.org~\cite{???} aids in the goal of distribution. PyPI hosts both
package source code (sdists) and binary artifacts (wheels). Anaconda.org
utilizes binary artifacts. Providing source code puts more of the onus of making
things work onto the end user, but also simplifies compatibility across
libraries. In general, it is too much to ask end users to set up compiler
toolchains to directly use source code. Providing a binary artifact requires
that the developer figures out build issues, but leaves the user with possible
binary compatibility problems with binary artifacts from other sources.

Unfortunately, preparing binary artifacts for distribution is often a bottleneck
as software developers are often domain experts but not necessarily expert in
compilation and distribution of software. Conda-Forge addresses this need, by
separating the code development and package management tasks. Conda-Forge is a
packaging ecosystem based around the conda package manager and anaconda.org
package repository. Conda-forge simplifies the developer burden by providing
standard build environment for all platforms (ensuring binary compatibility
across libraries) and by incorporating build and deployment into a streamlined
GithHub-based workflow. It focuses on being community-driven with high
centralized standards, and it serves as a gathering place for packaging
beginners and veterans alike. This paper describes the Conda Forge philosophy
and organization and we hope it will inspire people to join the conda-forge
community.

\section*{User's Perspective}

The user's goal is simple: get some kind of work done, and do as little
infrastructural work as possible to get that work done. As we discussed a bit
above, using binary packages is usually a good way to reduce that initial
infrastructural work, as dealing with compilers can be involved. Unfortunately,
using binaries doesn't always work as well as it should. There's a few reasons
why binaries break:

\begin{enumerate}
  \item Dependency libraries are an incompatible version
  \item Dependency libraries were compiled with an incompatible toolchain
  \item Library loading behavior of the operating system leads to usage of an
    incompatible dependency library
\end{enumerate}

% why use conda/CF
\item Conda is a binary package manager, etc..
\item compare and contrast conda with pypi
\item CF and conda:  from the user perspective, conda-forge is a channel on conda
\item Why should a user set conda-forge as default channel?
\begin{enumerate}
\item CF has many of the commonly used packages, and is growing every day
\item CF packages are guaranteed to interoperate with other CF packages
(which is a promise that not every channel can make)
\item it is easy to add packages to CF (if a user ever wanted to release their
own package)
\end{enumerate}
\item Setting up conda to use CF: Adding the CF channel to the \texttt{.condarc} to enable CF
inside conda
\end{enumerate}\seti

\section*{Package Maintainer's Perspective}
\begin{enumerate}\conti
\item Building packages for all platforms and architectures is hard, requiring
a lot of infrastructure and hardware
\item CF provides this infrastructure via continuous integration (CI) services,
allowing builds on Windows, Mac, and Linux platforms.
\item CF uses GitHub to hold the package recipes, the instructions for how to 
build the package, and the CI configuration files in repositories called ``feedstocks''
%\item probably need something about how package management is done in conda but outside of CF, so that when CF is described it looks much better than this.
\item Workflows:
\begin{enumerate}
    \item Submitting a new package
        \begin{enumerate}
            \item Fork \texttt{conda-forge/staged-recipes}
            \item Create a new branch
            \item Add the recipe in a new folder under \texttt{recipes}
            \item Open a pull request (PR) with the new recipe
            \item Once the PR has been merged, a new feedstock repo
                will be created and the folder removed
%\sjb{turn this around.  CF has a repo on GitHub called staged-recipes where package requirements are maintained.  New packages are added to staged-recipes via usual GitHub PR workflow.}
[view graph of staged-recipes workflow]
        \end{enumerate}
    \item Modifying an existing package
        \begin{enumerate}
            \item Fork the feedstock
            \item Create a new branch
            \item Change the existing recipe
            \item Open a pull request (PR) with the changes
            \item Once the PR has been merged, the package will be updated
[view graph of feedstock workflow]
        \end{enumerate}
\end{enumerate}
%\sjb{see above for the more gentle way of introducing these ideas}
\item Web services provide interactive tools for linting and modification of PRs
\end{enumerate}\seti

\section*{Ecosystem Maintainer's Perspective}
\begin{enumerate}\conti
\item The CF graph allows us to rebuild the entirety of conda-forge in
topological order [include view graph of graph]
\item One of the most common changes to packages is version bumps, which are
tedious to produce by hand, and were often out of sync with the source code
version changes, since recipe maintainers aren't necessarily aware of 
changes in all the packages they maintain.
\item The autotick bot was designed to address this issue by searching for
updates to the source code of all the packages that conda-forge manages,
and issuing PRs for packages which need to be updated.
\item Maintaining self consistency is especially important for binary packages.
Binary packages ship an Application Binary Interface (ABI) which describes how other software
can interact with the compiled binary.
\item Mixing ABIs can be disastrous as the mismatch in expected ABI and 
the actual ABI produces difficult to debug errors or erroneous results.
\item Without self consistency packages could depend on different ABIs
producing situations where packages are silently incompatible, forcing
users to choose which part of the ecosystem they can use.
\item To address this issue, and to make certain that each package in the 
ecosystem is compatible with the rest of the ecosystem CF has a global 
pinning repository.
This repository lists the current version of the package which CF builds
all of its packages against, effectively defining all the ABIs that CF ships.
\item The autotick bot infrastructure also helps maintain compatibility by
allowing for large scale migrations. These migrations move the entire ecosystem
from one ABI version to another, rebuilding all the packages in dependency
order.
\item The CF graph is accessible via \texttt{cf-graph3} and \texttt{libcfgraph}.
These repos allow users to query the graph and artifacts
\end{enumerate}\seti


\section*{Conclusions}
\begin{enumerate}\conti
\item CF provides a powerful platform for the distribution of scientific
software focusing on ecosystem stability, build traceability, and automation.
\item users: use it
\item software developers (and others): consider becoming CF contributors
\end{enumerate}\seti


\section*{Acknowledgments}

Work in the Billinge group was supported by U.S. Department of Energy,
Office of Science, Office of Basic Energy Sciences (DOE-BES) under contract
No. DE-SC00112704.

\bibliography{sample}

\end{document}
